---
title: "Manila Clam Survival Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options:
  chunk_output_type: console
---

This script analyzes survival assay data for lab-reared Manila clams in a factorial immune priming × heat priming experiment. Mortality (0 = alive, 1 = dead) is modeled over time using mixed-effects binomial logistic regression, with silo, field replicate, and assay plate included as random effects.

# Setup

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(lme4)
library(car)
library(emmeans)
library(DHARMa)
library(ggplot2)
library(cowplot)
library(lmerTest)
```

# Load data

```{r load}
# ── CONFIG ────────────────────────────────────────────────────────────────────
# Point this at your transformed mortality CSV (output of transform_mortality.R).
# If running multiple assay temperatures, either combine files before loading
# or re-run this script once per file.
DATA_FILE <- "data/survivalassay_SP1/WC_manila-40C_mortality.csv"
# ─────────────────────────────────────────────────────────────────────────────

surv <- read_csv(DATA_FILE)

# Confirm structure
glimpse(surv)
```

# Wrangle

Pivot timepoint columns to long format so each row is one clam × one timepoint.

```{r wrangle}
# Identify timepoint columns (anything ending in "HR")
tp_cols <- names(surv)[grepl("HR$", names(surv), ignore.case = TRUE)]

surv_long <- surv %>%
  pivot_longer(
    cols      = all_of(tp_cols),
    names_to  = "timepoint",
    values_to = "mortality"
  ) %>%
  mutate(
    # Parse numeric hours from labels like "4HR", "29.5HR"
    hours = as.numeric(sub("HR$", "", timepoint, ignore.case = TRUE)),

    # Ensure factor variables are correctly typed
    Immune_Priming    = factor(Immune_Priming,    levels = c("No", "Yes")),
    Heat_Priming      = factor(Heat_Priming,      levels = c("No", "Yes")),
    Silo_Replicate    = factor(Silo_Replicate),
    Field_Replicate   = factor(Field_Replicate),
    Assay_Replicate   = factor(Assay_Replicate),
    timepoint         = factor(timepoint, levels = tp_cols),  # preserve file order

    # Mortality must be integer 0/1 for binomial family
    mortality = as.integer(mortality)
  )

# Quick sanity check
surv_long %>%
  count(Immune_Priming, Heat_Priming, Silo_Replicate, Field_Replicate, Assay_Replicate) %>%
  print(n = Inf)
```

# Analysis: full time series

Mixed-effects binomial logistic regression across all timepoints.

Fixed effects: `Immune_Priming * Heat_Priming * hours`\
Random effects: `Silo_Replicate`, `Field_Replicate` (nested in silo), `Assay_Replicate` (plate)

*initial model – including silo, field rep, and assay rep as rand. effects*

```{r model-timeseries}
model_ts_full <- glmer(
  mortality ~ Immune_Priming * Heat_Priming * hours +
    (1 | Silo_Replicate) +
    (1 | Field_Replicate) +
    (1 | Assay_Replicate),
  family = binomial(link = "logit"),
  data   = surv_long,
  control = glmerControl(optimizer = "bobyqa",
                         optCtrl   = list(maxfun = 2e5))
)

summary(model_ts_full)
Anova(model_ts_full, type = "II")
```

Check significance of random effects.

```{r rand-effects-ts}
model_no_silo  <- update(model_ts_full, . ~ . - (1 | Silo_Replicate))
model_no_field <- update(model_ts_full, . ~ . - (1 | Field_Replicate))
model_no_assay <- update(model_ts_full, . ~ . - (1 | Assay_Replicate))

anova(model_ts_full, model_no_silo)
anova(model_ts_full, model_no_field)
anova(model_ts_full, model_no_assay)

VarCorr(model_ts_full)

## rewrite model without silo replicate and re run to determine if i should include silo:field or just 1 | field
```

```{r}
# initial model results on rand. effects: 
# 
### no silo vs. OG: 
# npar AIC BIC logLik -2*log(L) Chisq Df Pr(>Chisq) 
# model_no_silo 10 923.26 979.50 -451.63 903.26
# model_ts 11 925.26 987.13 -451.63 903.26 0 1 1
#
### no field vs. OG:
# npar    AIC     BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# model_no_field   10 943.93 1000.18 -461.97    923.93                         
# model_ts         11 925.26  987.13 -451.63    903.26 20.675  1  5.442e-06 ***
#
### no assay vs. OG:
# npar    AIC     BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# model_no_assay   10 955.90 1012.14 -467.95    935.90                         
# model_ts         11 925.26  987.13 -451.63    903.26 32.641  1  1.109e-08 ***
#
# will remove silo rand. effect from model and re run
```

```{r}
# re run model with silo rand. effect removed for model simplicity (& improved power?)

model_ts <- glmer(
  mortality ~ Immune_Priming * Heat_Priming * hours +
    (1 | Field_Replicate) +
    (1 | Assay_Replicate),
  family = binomial(link = "logit"),
  data   = surv_long,
  control = glmerControl(optimizer = "bobyqa",
                         optCtrl   = list(maxfun = 2e5))
)

summary(model_ts)
Anova(model_ts, type = "II")
```

Post-hoc: pairwise comparisons of treatment combinations at each timepoint.

```{r posthoc-ts}
emm_ts <- emmeans(model_ts, ~ Immune_Priming * Heat_Priming | hours, type = "response")
pairs(emm_ts)
```

### Testing as 4 independent treatments vs a 2x2 factorial.

```{r}
surv_long4 <- surv_long %>%
  mutate(Treatment = factor(case_when(
    Immune_Priming == "Yes" & Heat_Priming == "Yes" ~ "Combined",
    Immune_Priming == "Yes" & Heat_Priming == "No"  ~ "Immune_Only",
    Immune_Priming == "No"  & Heat_Priming == "Yes" ~ "Heat_Only",
    Immune_Priming == "No"  & Heat_Priming == "No"  ~ "Control"
  ), levels = c("Control", "Immune_Only", "Heat_Only", "Combined")))
```

```{r}
model_ts4 <- glmer(
  mortality ~ Treatment * hours +
    (1 | Field_Replicate) +
    (1 | Assay_Replicate),
  family = binomial(link = "logit"),
  data   = surv_long4,
  control = glmerControl(optimizer = "bobyqa",
                         optCtrl   = list(maxfun = 2e5))
)

summary(model_ts4)
Anova(model_ts4, type = "II")
```

```{r}
### testing for sig random effects
#
# model_no_silo4  <- update(model_ts4, . ~ . - (1 | Silo_Replicate))
# model_no_field4 <- update(model_ts4, . ~ . - (1 | Field_Replicate))
# model_no_assay4 <- update(model_ts4, . ~ . - (1 | Assay_Replicate))
# 
# anova(model_ts4, model_no_silo4)
# anova(model_ts4, model_no_field4)
# anova(model_ts4, model_no_assay4)
# 
# VarCorr(model_ts4)

###  only silo was non sig. rewrite model without silo replicate rand. effect term. 
```

```{r}
emm_ts4 <- emmeans(model_ts4, ~ Treatment | hours, type = "response")
pairs(emm_ts4)
```

## Write to .txt file

```{r}
output_file <- "output/WC_manila_S1/40C-model_output.txt"

tryCatch({
  sink(output_file, append = FALSE)
  
  cat("Run timestamp:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
  cat("================================================================\n\n")
  
  # # ── FULL FACTORIAL 2x2 MODELS ─────────────────────────────────────────────
  # cat("================================================================\n")
  # cat("2x2 FACTORIAL MODEL: Immune_Priming * Heat_Priming * hours\n")
  # cat("Random effects: Silo + Field + Assay (initial)\n")
  # cat("================================================================\n\n")
  # 
  # cat("--- Summary ---\n")
  # print(summary(model_ts_full))          # rename to whatever you called the 3-RE model
  # cat("\n--- Type II Anova ---\n")
  # print(Anova(model_ts_full, type = "II"))
  
  # # ── Random effects LRTs ───────────────────────────────────────────────────
  # cat("\n================================================================\n")
  # cat("RANDOM EFFECTS LIKELIHOOD RATIO TESTS (2x2 model)\n")
  # cat("================================================================\n\n")
  # 
  # cat("--- Silo removed vs. full ---\n");  print(anova(model_ts_full, model_no_silo))
  # cat("\n--- Field removed vs. full ---\n"); print(anova(model_ts_full, model_no_field))
  # cat("\n--- Assay removed vs. full ---\n"); print(anova(model_ts_full, model_no_assay))
  # cat("\n--- Variance components ---\n");    print(VarCorr(model_ts_full))
  
  # ── Simplified 2x2 model (silo removed) ───────────────────────────────────
  cat("\n================================================================\n")
  cat("2x2 FACTORIAL MODEL (SIMPLIFIED): Field + Assay random effects\n")
  cat("================================================================\n\n")
  
  cat("--- Summary ---\n")
  print(summary(model_ts))
  cat("\n--- Type II Anova ---\n")
  print(Anova(model_ts, type = "II"))
  
  # ── 2x2 Post-hoc ──────────────────────────────────────────────────────────
  cat("\n================================================================\n")
  cat("POST-HOC: 2x2 Pairwise comparisons at each timepoint\n")
  cat("================================================================\n\n")
  
  cat("--- EMmeans ---\n")
  print(emm_ts)
  cat("\n--- Pairwise contrasts ---\n")
  print(pairs(emm_ts))
  
  # ── 4-TREATMENT MODELS ────────────────────────────────────────────────────
  cat("\n================================================================\n")
  cat("4-TREATMENT MODEL: Treatment * hours\n")
  cat("Random effects: Field + Assay\n")
  cat("================================================================\n\n")
  
  cat("--- Summary ---\n")
  print(summary(model_ts4))
  cat("\n--- Type II Anova ---\n")
  print(Anova(model_ts4, type = "II"))
  
  # # ── 4-treatment random effects LRTs ───────────────────────────────────────
  # cat("\n================================================================\n")
  # cat("RANDOM EFFECTS LIKELIHOOD RATIO TESTS (4-treatment model)\n")
  # cat("================================================================\n\n")
  # 
  # cat("--- Field removed vs. full ---\n"); print(anova(model_ts4, model_no_field4))
  # cat("\n--- Assay removed vs. full ---\n"); print(anova(model_ts4, model_no_assay4))
  # cat("\n--- Variance components ---\n");    print(VarCorr(model_ts4))
  # 
  # ── 4-treatment Post-hoc ──────────────────────────────────────────────────
  cat("\n================================================================\n")
  cat("POST-HOC: 4-treatment pairwise comparisons at each timepoint\n")
  cat("================================================================\n\n")
  
  cat("--- EMmeans ---\n")
  print(emm_ts4)
  cat("\n--- Pairwise contrasts (Tukey) ---\n")
  print(pairs(emm_ts4, adjust = "tukey"))

}, finally = sink())

cat("Output written to:", output_file, "\n")
```

## Plot: predicted mortality over time

```{r plot-timeseries}
surv_long$predicted_mortality <- predict(model_ts, type = "response")

plot_ts <- ggplot(surv_long,
                  aes(x     = hours,
                      y     = predicted_mortality,
                      color = interaction(Immune_Priming, Heat_Priming),
                      fill  = interaction(Immune_Priming, Heat_Priming),
                      group = interaction(Immune_Priming, Heat_Priming))) +
  geom_point(aes(y = mortality), alpha = 0.2,
             position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = TRUE) +
  scale_colour_manual(
    values = c("No.No"   = "blue",
               "Yes.No"  = "green",
               "No.Yes"  = "red",
               "Yes.Yes" = "orange"),
    labels = c("No.No"   = "Control",
               "Yes.No"  = "Immune Priming with Poly(I:C)",
               "No.Yes"  = "Heat Priming at 33ºC",
               "Yes.Yes" = "Combined Immune + Heat Priming"),
    name = "Priming Treatment"
  ) +
  scale_fill_manual(
    values = c("No.No"   = "blue",
               "Yes.No"  = "green",
               "No.Yes"  = "red",
               "Yes.Yes" = "orange"),
    labels = c("No.No"   = "Control",
               "Yes.No"  = "Immune Priming with Poly(I:C)",
               "No.Yes"  = "Heat Priming at 33ºC",
               "Yes.Yes" = "Combined Immune + Heat Priming"),
    name = "Priming Treatment"
  ) +
  labs(
    title = paste("Predicted Mortality over Time —", unique(surv_long$Assay_Temp)),
    x     = "Hours",
    y     = "Probability of Mortality"
  ) +
  theme_classic(base_size = 13); plot_ts

ggsave(plot_ts,
       filename = "output/WC_manila_S1/WC40C-survival_timeseries.png",
       width = 8, height = 5)
```

### Plot individual plates (assay replicates)

```{r}
plot3 <-ggplot(surv_long, 
              aes(x = hours, 
                  y = predicted_mortality, 
                  color = Assay_Replicate, 
                  fill = Assay_Replicate, 
                  group = Assay_Replicate)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  scale_colour_manual(
    values = c("WC-CC1-40"   = "blue",
               "WC-CC2-40"  = "darkblue",
               "WC-CP1-40"  = "forestgreen",
               "WC-CP2-40" = "darkgreen",
               "WC-EE1-40"   = "red",
               "WC-EE2-40"  = "darkred",
               "WC-EP1-40"  = "orange",
               "WC-EP2-40" = "goldenrod"),

    labels = c("WC-CC1-40"   = "Control 1",
               "WC-CC2-40"  = "Control 2",
               "WC-CP1-40"  = "Immune 1",
               "WC-CP2-40" = "Immune 2",
               "WC-EE1-40"   = "Heat 1",
               "WC-EE2-40"  = "Heat 2",
               "WC-EP1-40"  = "Combined 1",
               "WC-EP2-40" = "Combined 2"),
    name = "Priming Treatment"
  ) +
  
  scale_fill_manual(
    values = c("WC-CC1-40"   = "blue",
               "WC-CC2-40"  = "darkblue",
               "WC-CP1-40"  = "forestgreen",
               "WC-CP2-40" = "darkgreen",
               "WC-EE1-40"   = "red",
               "WC-EE2-40"  = "darkred",
               "WC-EP1-40"  = "orange",
               "WC-EP2-40" = "goldenrod"),

    labels = c("WC-CC1-40"   = "Control 1",
               "WC-CC2-40"  = "Control 2",
               "WC-CP1-40"  = "Immune 1",
               "WC-CP2-40" = "Immune 2",
               "WC-EE1-40"   = "Heat 1",
               "WC-EE2-40"  = "Heat 2",
               "WC-EP1-40"  = "Combined 1",
               "WC-EP2-40" = "Combined 2"),
    name = "Priming Treatment"
  ) +
  
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  
  theme_classic();plot3
```

## KM Curves

## Generate Kaplan Meier survival curves

Plot the survival function

```{r}
library(survival)
library(ggsurvfit)
library(gtsummary)
```

```{r}
plot_km <- survfit2(
  Surv(hours, mortality) ~ Immune_Priming + Heat_Priming,
  data = surv_long
) %>%
  ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability"
  )

plot_km
```

Use a log rank model to determine statistical differences in curves.

```{r}
survival::survdiff(Surv(hours, mortality) ~ Immune_Priming + Heat_Priming,
  data = surv_long)
```

Analyze again with a Cox proportional hazards model.

```{r}
survival::coxph(Surv(hours, mortality) ~ Immune_Priming*Heat_Priming, data = surv_long)

survival::coxph(Surv(hours, mortality) ~ Immune_Priming*Heat_Priming, data = surv_long) %>% 
  tbl_regression(exp = TRUE) 
```

## Generate plots (still working)

Set theme.

```{r}
my_theme<-theme_classic()+
  theme(
    axis.title.y = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size=12, color = "black"),
    axis.text.x = element_text(size=12, color = "black"),
    axis.title.x =element_text(size = 14, color = "black"),
    legend.position = "none")
```

Stress treatment

```{r}
#options("ggsurvfit.switch-color-linetype" = TRUE)
plot1<-ggsurvfit::survfit2(Surv(hours, mortality) ~ Immune_Priming * Heat_Priming, data = surv_long) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  #scale_linetype_manual(values=c("solid", "dashed"))+
  scale_colour_manual(values=c("darkgray", "orange"))+
  theme(legend.position="right");plot1
```

Stress treatment

```{r}
#options("ggsurvfit.switch-color-linetype" = TRUE)
plot2<-ggsurvfit::survfit2(Surv(time, status) ~ Family, data = data) %>% 
  ggsurvfit::ggsurvfit() +
  labs(
    x = "Hours",
    y = "Survival probability",
    title=""
  )+
  ylim(0,1)+
  my_theme+
  theme(legend.position="right");plot2
```
