---
title: APC cockle resazurin analysis
author: "Sam White"
date: '2026'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

This script analyzes resazurin assays for APC cockle spat. Treatments
are decoded as a 2x2 factorial: Immune_Priming (Yes/No) x Heat_Priming
(Yes/No), derived from treatment codes CC, EE, CP, EP. Metadata is
provided via layout.csv.

Treatment key: CC = Immune No / Heat No (control) EE = Immune No / Heat
Yes (heat priming only) CP = Immune Yes / Heat No (immune priming only)
EP = Immune Yes / Heat Yes (combined)

# Set up

Set up workspace, set options, and load required packages.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)   # data wrangling and plotting
library(MASS)        # statistical functions (load before tidyverse to avoid select() conflict)
select <- dplyr::select  # restore dplyr::select after MASS masks it
library(ggplot2)     # plotting
library(readxl)      # reading Excel files
library(cowplot)     # plot composition
library(lme4)        # mixed effects models
library(lmerTest)    # p-values for lmer via Satterthwaite approximation
library(car)         # Type II/III ANOVA for mixed models
library(effects)     # model effect plots
library(emmeans)     # estimated marginal means and pairwise contrasts
library(tools)       # file utility functions (file_path_sans_ext)
library(pracma)      # trapz() for AUC calculation
```

# Configuration

Specify input data and output directory paths here. Update these before
running.

```{r config}
# ── INPUT ─────────────────────────────────────────────────────────────────────
# Folder containing APC plate .txt files and layout.csv
data_dir <- "data/resassay_SP1/AP_cockle_final"

# ── OUTPUT ────────────────────────────────────────────────────────────────────
# Directory for saved figures
fig_dir  <- "output/AP_cockles_SP1_pink/figures/ah"

# Create output directories if they don't already exist
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

# Treatment lookup table

Maps 2-letter treatment codes to 2x2 factorial factor levels. Used
downstream to decode treatment_group into Immune_Priming and
Heat_Priming columns.

```{r treatment-lookup}
TREATMENT_INFO <- list(
  "CC" = c(Immune_Priming = "No",  Heat_Priming = "No"),   # control
  "EE" = c(Immune_Priming = "No",  Heat_Priming = "Yes"),  # heat priming only
  "CP" = c(Immune_Priming = "Yes", Heat_Priming = "No"),   # immune priming only
  "EP" = c(Immune_Priming = "Yes", Heat_Priming = "Yes")   # combined
)

# Shared colour palette for the 4 treatment combinations; used in all plots
trt_colours <- c(
  "Control"      = "blue4",       # CC
  "Immune Only"  = "forestgreen", # CP
  "Heat Only"    = "red",      # EE
  "Combined"     = "orange"      # EP
)
```

# Resazurin metabolic rates (individual)

## Load data

```{r load-plates}
# Match only APC-***-##-T#.txt files; excludes layout.csv and other files
file_list <- list.files(
  path = data_dir,
  pattern = "^APC-[A-Z]+[0-9]+-[0-9]+-T[0-9]+\\.txt$",
  full.names = TRUE
)

# Stop early if no files found rather than silently producing empty output
if (length(file_list) == 0) {
  stop("No matching txt files found in the specified folder. Check the folder path.")
}

data_list <- list()  # accumulate parsed data frames here

for (file in file_list) {
  if (!is.character(file)) next  # skip non-character entries (safety check)

  file_name <- file_path_sans_ext(basename(file))  # e.g. "APC-CC1-26-T0"

  # Scan for the "Results" section header in the plate reader export
  lines       <- readLines(file, warn = FALSE)
  results_idx <- which(trimws(lines) == "Results")
  if (length(results_idx) == 0) next  # skip files with no Results block
  idx <- results_idx[1]               # use first Results block if multiple

  # Read the tab-delimited data starting at the Results block
  data <- read.delim(file, header = FALSE, skip = idx,
                     stringsAsFactors = FALSE, fill = TRUE)

  # Promote the first data row (column numbers) to column names
  col_header    <- as.character(unlist(data[1, ]))
  col_header[1] <- "Row"           # first cell is blank; label it "Row"
  colnames(data) <- col_header
  data <- data[-1, ]               # drop the now-redundant header row
  data <- data[, colnames(data) != ""]  # drop empty trailing columns

  # Keep only true plate rows (single letters A–D for a 24-well plate)
  data <- data %>% filter(str_detect(Row, "^[A-Za-z]$"))

  # Pivot to long format: one row per well per file
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Value      = suppressWarnings(as.numeric(Value)),  # coerce to numeric; non-numeric -> NA
      plate_well = paste0(toupper(Row), Column),         # well ID e.g. "A2", "B3"
      FileName   = file_name,
      plate_ID   = str_remove(file_name, "-T[0-9]+$"),  # strip timepoint suffix e.g. "APC-CC1-26"
      timepoint  = str_extract(file_name, "T([0-9]+)$") %>%
        str_remove("T") %>% as.numeric()                # extract numeric timepoint
    ) %>%
    filter(!is.na(Value)) %>%                           # drop wells with no reading
    select(FileName, plate_well, Value, plate_ID, timepoint)

  data_list[[file_name]] <- data_long
}

head(data_list[[file_name]])  # preview last parsed file

# Stack all per-file data frames into one
combined_data <- bind_rows(data_list, .id = "Source")

combined_data <- combined_data %>%
  rename(resazurin_counts = Value) %>%       # rename for clarity
  mutate(timepoint = as.character(timepoint)) # keep timepoint as character for grouping

head(combined_data)
```

## Load metadata

```{r load-metadata}
# Read layout.csv; treat all columns as character initially for safe parsing
metadata <- read_csv(file.path(data_dir, "layout.csv"),
                     col_types = cols(.default = "c"),
                     show_col_types = FALSE) %>%
  mutate(
    # Parse is_blank: accept TRUE/T/1/YES/Y (case-insensitive) as TRUE
    is_blank = toupper(trimws(is_blank)) %in% c("TRUE", "T", "1", "YES", "Y"),
    # Parse exclude_from_analysis the same way
    exclude_from_analysis = toupper(trimws(exclude_from_analysis)) %in%
      c("TRUE", "T", "1", "YES", "Y"),
    temperature = as.character(temperature),
    # Derive a type column to match the original script's "blank"/"sample" logic
    type = if_else(is_blank, "blank", "sample")
  )
```

## Join plate data with metadata

```{r join-data}
str(combined_data)
str(metadata)

# Left join: every plate well gets its metadata; wells with no layout entry are dropped
full_data <- left_join(combined_data, metadata,
                       by = c("plate_ID", "plate_well")) %>%
  filter(!is.na(type))  # drop wells not present in layout

head(full_data)
```

## Add size and exclude flagged individuals

```{r size-and-exclusions}
# area_mm2.measurement from layout.csv is the size proxy; rename for downstream compatibility
full_data <- full_data %>%
  mutate(length.mm = suppressWarnings(as.numeric(area_mm2.measurement)))

# Report excluded wells before dropping them
excluded_wells <- full_data %>%
  filter(exclude_from_analysis == TRUE) %>%
  mutate(
    # Use sample_id_group as label if present; otherwise fall back to plate+well
    sample = if_else(
      !is.na(sample_id_group) & trimws(as.character(sample_id_group)) != "",
      as.character(sample_id_group),
      paste(plate_ID, plate_well, sep = "_")
    )
  ) %>%
  select(plate_ID, plate_well, sample, fieldrep_ID, treatment_group,
         any_of("exclude_reason")) %>%
  distinct() %>%
  arrange(plate_ID, plate_well)
```

```{r excluded-wells, results='asis'}
if (nrow(excluded_wells) > 0) {
  # Build col_names dynamically so count always matches ncol(excluded_wells)
  col_map <- c(
    plate_ID        = "Plate",
    plate_well      = "Well",
    sample          = "Sample",
    fieldrep_ID     = "Family",
    treatment_group = "Treatment",
    exclude_reason  = "Reason"
  )
  col_names <- col_map[names(col_map) %in% names(excluded_wells)]
  cat(knitr::kable(excluded_wells, col.names = col_names), sep = "\n")
} else {
  cat("No wells are excluded from analysis.\n")
}
```

```{r apply-exclusions}
# Remove excluded individuals from all downstream calculations
full_data <- full_data %>%
  filter(is.na(exclude_from_analysis) | exclude_from_analysis == FALSE)
```

## Decode 2x2 factorial treatment structure

```{r factorial-decode}
# Decode treatment_group (CC/EE/CP/EP) into two binary factors.
# This is the key step that enables Immune_Priming * Heat_Priming models.
full_data <- full_data %>%
  mutate(
    Immune_Priming = factor(
      sapply(toupper(treatment_group), function(trt) {
        info <- TREATMENT_INFO[[trt]]
        if (is.null(info)) NA_character_ else info["Immune_Priming"]
      }),
      levels = c("No", "Yes")  # "No" as reference level
    ),
    Heat_Priming = factor(
      sapply(toupper(treatment_group), function(trt) {
        info <- TREATMENT_INFO[[trt]]
        if (is.null(info)) NA_character_ else info["Heat_Priming"]
      }),
      levels = c("No", "Yes")  # "No" as reference level
    ),
    # Human-readable treatment label for plots
    Treatment = case_when(
      Immune_Priming == "No"  & Heat_Priming == "No"  ~ "Control",
      Immune_Priming == "Yes" & Heat_Priming == "No"  ~ "Immune Only",
      Immune_Priming == "No"  & Heat_Priming == "Yes" ~ "Heat Only",
      Immune_Priming == "Yes" & Heat_Priming == "Yes" ~ "Combined"
    )
  )

# Confirm decoding worked
table(full_data$treatment_group, full_data$Treatment, useNA = "always")
```

## Prep the data

```{r plot-raw}
# Raw fluorescence by plate — colour by Treatment to check plate layout visually
full_data %>%
  ggplot(aes(x = timepoint, y = resazurin_counts,
             colour = Treatment,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = trt_colours) +
  theme_classic()
```

```{r fold-change}
# Normalise each well's fluorescence to its own T0 value (fold-change from baseline).
# group_by ensures the first() call is taken within each individual well's time series.
full_data <- full_data %>%
  group_by(plate_ID, plate_well, temperature, treatment_group,
           Immune_Priming, Heat_Priming, Treatment, fieldrep_ID, plate_rep) %>%
  arrange(plate_ID, plate_well, as.numeric(timepoint)) %>%
  mutate(fluorescence.norm = resazurin_counts / first(resazurin_counts))
```

```{r plot-norm}
# Fold-change traces by plate — quick check for outlier wells or bad timepoints
full_data %>%
  ggplot(aes(x = timepoint, y = fluorescence.norm,
             colour = Treatment,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = trt_colours) +
  theme_classic()
```

```{r view-blanks}
# Inspect blank well traces to assess background drift before subtraction
full_data %>%
  filter(type == "blank") %>%
  ggplot(aes(x = timepoint, y = fluorescence.norm,
             colour = plate_ID,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

```{r blank-means}
# Compute mean blank fold-change per plate per timepoint.
# This is the background correction value that will be subtracted from each sample.
blanks <- full_data %>%
  filter(type == "blank") %>%
  group_by(plate_ID, temperature, timepoint) %>%
  summarise(mean_blank = mean(fluorescence.norm, na.rm = TRUE), .groups = "drop")
blanks
```

```{r plot-blank-means}
# Plot mean blank by plate — should be close to 1.0 with minimal drift
blanks %>%
  ggplot(aes(x = timepoint, y = mean_blank, colour = plate_ID, group = plate_ID)) +
  geom_point() +
  geom_line() +
  theme_classic()
```

```{r blank-subtract}
# Join per-plate blank means onto full_data, then subtract from sample fold-change.
# Removes background fluorescence drift; all wells start at ~0 at T0 by construction.
full_data <- left_join(full_data, blanks, by = c("plate_ID", "temperature", "timepoint"))

full_data <- full_data %>%
  filter(type != "blank") %>%           # drop blank wells from analysis data
  mutate(value = fluorescence.norm - mean_blank)  # blank-corrected fold-change
```

```{r plot-corrected}
# Blank-corrected fold-change by plate — check that correction looks reasonable
full_data %>%
  ggplot(aes(x = timepoint, y = value,
             colour = Treatment,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = trt_colours) +
  theme_classic()
```

```{r size-normalise}
# Divide blank-corrected fold-change by individual area (mm²) to get
# size-normalised metabolic rate (value.mm = fold-change per mm²)
full_data <- full_data %>%
  mutate(value.mm = value / length.mm)
```

```{r plot-size-normalised}
# Size-normalised metabolic rate by plate
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = Treatment,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = trt_colours) +
  theme_classic()
```

```{r remove-cols}
# Drop intermediate columns no longer needed to keep full_data tidy
full_data <- full_data %>%
  select(!resazurin_counts) %>%  # raw counts replaced by normalised values
  select(!FileName) %>%          # file provenance no longer needed
  select(!type) %>%              # blank/sample flag no longer needed
  select(!fluorescence.norm) %>% # intermediate normalisation step
  select(!mean_blank)            # blank correction value no longer needed
```

## Models

```{r plot-by-treatment}
# Metabolic rate traces faceted by Treatment — visual check before modelling
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = temperature,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ Treatment) +
  geom_point() +
  geom_line() +
  theme_classic()
```

```{r plot-by-temp}
# Metabolic rate traces faceted by temperature — colour by Treatment
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = Treatment,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ temperature) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = trt_colours) +
  theme_classic()
```

```{r model-initial}
# Initial LME: 2x2 factorial fixed effects (Immune x Heat x timepoint).
# Random effects: plate_ID:plate_well accounts for repeated measures on the same well
# across timepoints; fieldrep_ID accounts for family-level variation.
model <- lmer(value.mm ~ timepoint * Immune_Priming * Heat_Priming +
                (1 | plate_ID:plate_well) +
                (1 | fieldrep_ID),
              data = full_data)

summary(model)
car::Anova(model, type = "II")  # Type II ANOVA: tests main effects + interactions
qqPlot(residuals(model))        # check normality of residuals
```

```{r model-cuberoot}
# Residuals often non-normal for fluorescence data; cube-root transformation
# (1 + value.mm)^(1/3) compresses positive skew while keeping values >= 0.
model <- lmer((1 + value.mm)^(1/3) ~ timepoint * Immune_Priming * Heat_Priming +
                (1 | plate_ID:plate_well) +
                (1 | fieldrep_ID),
              data = full_data)

summary(model)
anova(model)          # Satterthwaite F-tests from lmerTest
qqPlot(residuals(model))
```

```{r posthoc-timeseries}
# Pairwise contrasts for the 2x2 factorial at each timepoint.
# Tukey adjustment controls family-wise error rate across all comparisons.

# Both factors combined: all 4 treatment combinations vs each other at each timepoint
emm <- emmeans(model, ~ Immune_Priming * Heat_Priming | timepoint)
pairs(emm, adjust = "tukey")

# Main effect of immune priming (collapsed across heat priming levels) at each timepoint
emm <- emmeans(model, ~ Immune_Priming | timepoint)
pairs(emm, adjust = "tukey")

# Main effect of heat priming (collapsed across immune priming levels) at each timepoint
emm <- emmeans(model, ~ Heat_Priming | timepoint)
pairs(emm, adjust = "tukey")

# Temperature effect at each timepoint (not a treatment factor — for reference only)
emm <- emmeans(model, ~ temperature | timepoint)
pairs(emm, adjust = "tukey")
```

## Plotting

```{r plot1-by-family}
# Metabolic rate by family (fieldrep_ID) and plate replicate — useful for
# diagnosing family-level or plate-level effects before interpreting treatment effects
plot1 <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = Treatment, fill = Treatment)) +
  facet_grid(~ fieldrep_ID * plate_rep) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = Treatment)) +
  scale_colour_manual(values = trt_colours) +
  scale_fill_manual(values = trt_colours) +
  theme_classic() +
  xlab("Hour")
plot1
```

```{r plot1a-treatment-overall}
# Overall metabolic rate by Treatment across all plates and temperatures
# loess smoother shows the average trajectory per treatment group
plot1a <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = Treatment, fill = Treatment, group = Treatment)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "loess") +
  scale_colour_manual(values = trt_colours) +
  scale_fill_manual(values = trt_colours) +
  theme_classic() +
  xlab("Hour")
plot1a

ggsave(plot1a, filename = file.path(fig_dir, "metabolism_treatment.png"), width = 7, height = 5)
```

```{r plot1b-by-family-temp}
# Metabolic rate by family (fieldrep_ID) faceted by temperature —
# helps visualise whether family effects are consistent across temperatures
plot1b <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = fieldrep_ID, fill = fieldrep_ID)) +
  facet_grid(~ temperature) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = fieldrep_ID)) +
  theme_classic() +
  xlab("Hour")
plot1b
```

```{r plot1c-treatment-by-temp}
# Treatment effect on metabolic rate at each temperature — key visualisation
# for the Immune_Priming x Heat_Priming x temperature interaction
plot1c <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = Treatment, fill = Treatment)) +
  facet_grid(~ temperature) +
  geom_smooth(aes(group = Treatment)) +
  scale_colour_manual(values = trt_colours, name = "Priming Treatment") +
  scale_fill_manual(values = trt_colours, name = "Priming Treatment") +
  theme_classic() +
  theme(axis.text.y  = element_text(size = 17),
        axis.text.x  = element_text(size = 17),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17)) +
  ylab("Metabolic Rate") +
  xlab("Hour")
plot1c

ggsave(plot1c, filename = file.path(fig_dir, "metabolism_treatment_by_temp.png"), width = 7, height = 4)
```

```{r plot1d-temp-by-treatment}
# Temperature effect on metabolic rate within each treatment —
# complements plot1c by swapping the faceting and colour variables
plot1d <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = temperature, fill = temperature, group = temperature)) +
  facet_grid(~ Treatment) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "loess") +
  theme_classic() +
  xlab("Hour")
plot1d

ggsave(plot1d, filename = file.path(fig_dir, "metabolism_temperature_by_treatment.png"), width = 8, height = 5)
```

# Survival

```{r survival-stub}
# NOTE: No separate survival file for this dataset.
# To enable survival analysis: add a 'mortality' column (0 = alive, 1 = dead)
# to layout.csv, then model with:
#   glmer(mortality ~ temperature * Immune_Priming * Heat_Priming +
#           (1 | plate_rep) + (1 | fieldrep_ID),
#         family = binomial(link = "logit"), data = surv_model_data)
message("Survival data not available in layout.csv; skipping survival sections.")
```

# Metabolic rates across thermal range (AUC)

AUC summarises total metabolic activity per individual across all
timepoints. A larger AUC = more cumulative metabolic activity over the
assay period.

```{r auc-calc}
# Compute trapezoidal AUC per individual (plate_ID x plate_well).
# Carry Immune_Priming, Heat_Priming, and Treatment through for modelling and plotting.
auc_data <- full_data %>%
  mutate(time = as.numeric(timepoint)) %>%
  group_by(plate_ID, plate_well, temperature, treatment_group,
           Immune_Priming, Heat_Priming, Treatment, fieldrep_ID, plate_rep) %>%
  summarise(AUC = trapz(time, value.mm), .groups = "drop") %>%  # trapezoid rule
  mutate(temperature_num = as.numeric(temperature))  # numeric temperature for ordering

head(auc_data)
```

```{r auc-hist}
hist(auc_data$AUC)  # check distribution; heavy skew may warrant log transformation
```

```{r plot5-auc-raw}
# Raw AUC points by Treatment and temperature — preliminary look before modelling
plot5 <- ggplot(auc_data, aes(x = temperature, y = AUC,
                              color = Treatment, fill = Treatment,
                              group = Treatment)) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01)) +
  stat_smooth(aes(group = Treatment)) +
  scale_colour_manual(values = trt_colours, name = "Priming Treatment") +
  scale_fill_manual(values = trt_colours, name = "Priming Treatment") +
  labs(title = "A", y = "AUC", x = "Temperature") +
  theme_classic(base_size = 16)
plot5
```

```{r auc-model}
# 2x2 factorial AUC model: log(1+AUC) to address positive skew.
# Fixed: temperature * Immune_Priming * Heat_Priming (full factorial + temperature interaction).
# Random: fieldrep_ID (family), plate_ID (assay plate variation).
tpc_model <- lmer(log(1 + AUC) ~ temperature * Immune_Priming * Heat_Priming +
                    (1 | fieldrep_ID) +
                    (1 | plate_ID),
                  data = auc_data)

summary(tpc_model)
anova(tpc_model, type = "II")  # Type II ANOVA from lmerTest
qqPlot(residuals(tpc_model))   # normality check
plot(tpc_model, which = 1)     # residuals vs fitted (check homoscedasticity)
plot(tpc_model, which = 3)     # scale-location plot
```

```{r auc-posthoc}
# Pairwise contrasts for all 4 treatment combinations at each temperature
emm <- emmeans(tpc_model, ~ Immune_Priming * Heat_Priming | temperature)
pairs(emm, adjust = "tukey")

# Main effect of immune priming at each temperature (collapsed across heat priming)
emm <- emmeans(tpc_model, ~ Immune_Priming | temperature)
pairs(emm, adjust = "tukey")

# Main effect of heat priming at each temperature (collapsed across immune priming)
emm <- emmeans(tpc_model, ~ Heat_Priming | temperature)
pairs(emm, adjust = "tukey")
```

```{r auc-pred-grid}
# Build a prediction grid covering all temperature x Immune_Priming x Heat_Priming
# combinations; set random effects to NA so predictions are marginal (population-level)
pred_grid <- expand.grid(
  temperature    = levels(as.factor(auc_data$temperature)),
  Immune_Priming = levels(auc_data$Immune_Priming),
  Heat_Priming   = levels(auc_data$Heat_Priming),
  fieldrep_ID    = NA   # marginal prediction; ignore random effect
) %>%
  mutate(
    temperature    = factor(temperature, levels = levels(as.factor(auc_data$temperature))),
    Immune_Priming = factor(Immune_Priming, levels = c("No", "Yes")),
    Heat_Priming   = factor(Heat_Priming,   levels = c("No", "Yes")),
    # Add Treatment label for colour mapping in the plot
    Treatment = case_when(
      Immune_Priming == "No"  & Heat_Priming == "No"  ~ "Control",
      Immune_Priming == "Yes" & Heat_Priming == "No"  ~ "Immune Only",
      Immune_Priming == "No"  & Heat_Priming == "Yes" ~ "Heat Only",
      Immune_Priming == "Yes" & Heat_Priming == "Yes" ~ "Combined"
    )
  )

# Predict on log scale (re.form = NA = marginal/population-level prediction)
pred_grid$logAUC_pred <- predict(tpc_model, newdata = pred_grid, re.form = NA)

# Back-transform from log(1+AUC) to AUC scale
pred_grid$AUC_pred <- exp(pred_grid$logAUC_pred) - 1
```

```{r plot5-auc-model}
# Model prediction plot: raw data as jittered points, model predictions as lines.
# Separates the 4 treatment combinations by colour using the shared trt_colours palette.
plot5 <- ggplot() +
  # Raw individual AUC values (semi-transparent to show density)
  geom_point(data = auc_data,
             aes(x = temperature, y = AUC, color = Treatment),
             alpha = 0.3, position = position_jitter(height = 0.01)) +
  # Model-predicted AUC trajectory per treatment across temperatures
  geom_line(data = pred_grid,
            aes(x = temperature, y = AUC_pred,
                color = Treatment, group = Treatment),
            linewidth = 1.5) +
  scale_colour_manual(values = trt_colours, name = "Priming Treatment") +
  labs(title = "A", y = "AUC", x = "Temperature") +
  theme_classic(base_size = 16)
plot5

ggsave(plot5, filename = file.path(fig_dir, "auc-temperatures.png"), width = 7, height = 5)
```

```{r auc-summary}
# Group-level mean AUC summary table: one column per treatment, one row per temperature.
# Useful for reporting and calculating percent differences between groups.
auc_data %>%
  group_by(temperature, Treatment) %>%
  summarise(mean = mean(AUC, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Treatment, values_from = mean)
```
