---
title: APC cockle resazurin analysis
author: "Sam White"
date: '2026'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays for APC cockle spat. Metadata is provided via layout.csv with columns: plate_ID, plate_rep, temperature, plate_well, is_blank, area_mm2.measurement, imageJ_ID, fieldrep_ID, treatment_group, sample_id_group, exclude_from_analysis, exclude_reason.

# Set up

Set up workspace, set options, and load required packages.

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries.

```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
library(tools)
```

# Resazurin metabolic rates (individual)

## Load data

```{r}
# Set the folder path — update to your plate file directory
folder_path <- "data/resassay_SP1/AP_cockle_final/"

# List all APC-*-T*.txt files in the folder
file_list <- list.files(
  path = folder_path,
  pattern = "^APC-[A-Z]+[0-9]+-[0-9]+-T[0-9]+\\.txt$",
  full.names = TRUE
)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No matching txt files found in the specified folder. Check the folder path.")
}

# Initialize an empty list to store processed data
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  if (!is.character(file)) next

  file_name <- file_path_sans_ext(basename(file))

  # Find the Results block
  lines <- readLines(file, warn = FALSE)
  results_idx <- which(trimws(lines) == "Results")
  if (length(results_idx) == 0) next
  idx <- results_idx[1]

  # Read data starting from the header row of the Results block
  data <- read.delim(file, header = FALSE, skip = idx,
                     stringsAsFactors = FALSE, fill = TRUE)

  # First row is the column header (blank + column numbers)
  col_header <- as.character(unlist(data[1, ]))
  col_header[1] <- "Row"
  colnames(data) <- col_header
  data <- data[-1, ]   # drop header row
  data <- data[, colnames(data) != ""]  # drop empty trailing columns

  # Keep only rows that start with a letter (plate rows A–D)
  data <- data %>% filter(str_detect(Row, "^[A-Za-z]$"))

  # Convert to long format
  data_long <- data %>%
    pivot_longer(cols = -Row, names_to = "Column", values_to = "Value") %>%
    mutate(
      Value    = suppressWarnings(as.numeric(Value)),
      plate_well = paste0(toupper(Row), Column),   # e.g. "A2", "B3"
      FileName = file_name,
      # plate_ID is everything before -T<n>.txt
      plate_ID  = str_remove(file_name, "-T[0-9]+$"),
      timepoint = str_extract(file_name, "T([0-9]+)$") %>%
        str_remove("T") %>% as.numeric()
    ) %>%
    filter(!is.na(Value)) %>%
    dplyr::select(FileName, plate_well, Value, plate_ID, timepoint)

  data_list[[file_name]] <- data_long
}

head(data_list[[file_name]])

combined_data <- bind_rows(data_list, .id = "Source")

# Rename value column
combined_data <- combined_data %>%
  rename(resazurin_counts = Value) %>%
  mutate(timepoint = as.character(timepoint))

head(combined_data)
```

Load in metadata.

```{r}
metadata <- read_csv("data/resassay_SP1/AP_cockle_final/layout.csv",
                     col_types = cols(.default = "c"),
                     show_col_types = FALSE) %>%
  mutate(
    is_blank      = toupper(trimws(is_blank)) %in% c("TRUE", "T", "1", "YES", "Y"),
    exclude_from_analysis = toupper(trimws(exclude_from_analysis)) %in%
      c("TRUE", "T", "1", "YES", "Y"),
    temperature   = as.character(temperature),
    # type column: mirrors original script's "blank"/"sample" logic
    type          = if_else(is_blank, "blank", "sample")
  )
```

Join with data frame and remove any wells that did not have samples.

```{r}
str(combined_data)
str(metadata)

full_data <- left_join(combined_data, metadata,
                       by = c("plate_ID", "plate_well")) %>%
  filter(!is.na(type))

head(full_data)
```

Load in size data.

```{r}
# Size data (area_mm2) is already in the layout/metadata join.
# Rename area_mm2.measurement to length.mm for compatibility with downstream code.
full_data <- full_data %>%
  mutate(length.mm = suppressWarnings(as.numeric(area_mm2.measurement)))
```

<!-- Size (area_mm2) already joined from layout above. -->

## Prep the data

Plot the raw data.

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = resazurin_counts,
             colour = temperature,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

Calculate fluorescence at each time point normalized to the starting value at time 0.

```{r}
full_data <- full_data %>%
  group_by(plate_ID, plate_well, temperature, treatment_group, fieldrep_ID, plate_rep) %>%
  arrange(plate_ID, plate_well, as.numeric(timepoint)) %>%
  mutate(fluorescence.norm = resazurin_counts / first(resazurin_counts))
```

Plot again.

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = fluorescence.norm,
             colour = treatment_group,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

View blanks

```{r}
full_data %>%
  filter(type == "blank") %>%
  ggplot(aes(x = timepoint, y = fluorescence.norm,
             colour = plate_ID,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

Calculate mean change in blank at each time point.

```{r}
blanks <- full_data %>%
  filter(type == "blank") %>%
  group_by(plate_ID, temperature, timepoint) %>%
  summarise(mean_blank = mean(fluorescence.norm, na.rm = TRUE), .groups = "drop")
blanks
```

View summarized blank data.

```{r}
blanks %>%
  ggplot(aes(x = timepoint, y = mean_blank, colour = plate_ID, group = plate_ID)) +
  geom_point() +
  geom_line() +
  theme_classic()
```

Subtract blank values from fluorescence values for oysters.

```{r}
full_data <- left_join(full_data, blanks, by = c("plate_ID", "temperature", "timepoint"))

full_data <- full_data %>%
  filter(type != "blank") %>%
  mutate(value = fluorescence.norm - mean_blank)
```

Plot again.

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = value,
             colour = treatment_group,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

Size normalize data.

```{r}
full_data<-full_data%>%
  mutate(value.mm=value/length.mm)
```

Plot again.

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = treatment_group,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ plate_ID) +
  geom_point() +
  geom_line() +
  theme_classic()
```

Remove unnecessary columns.

```{r}
full_data <- full_data %>%
  select(!resazurin_counts) %>%
  select(!FileName) %>%
  select(!type) %>%
  select(!fluorescence.norm) %>%
  select(!mean_blank)
```

## Excluded wells

```{r excluded-wells, results='asis'}
excluded_wells <- full_data %>%
  filter(exclude_from_analysis == TRUE) %>%
  mutate(
    sample = if_else(
      !is.na(sample_id_group) & trimws(as.character(sample_id_group)) != "",
      as.character(sample_id_group),
      paste(plate_ID, plate_well, sep = "_")
    )
  ) %>%
  dplyr::select(plate_ID, plate_well, sample, fieldrep_ID, treatment_group,
                any_of("exclude_reason")) %>%
  distinct() %>%
  arrange(plate_ID, plate_well)

if (nrow(excluded_wells) > 0) {
  col_map <- c(
    temperature     = "Temperature",
    plate_rep       = "Plate Rep",
    plate_ID        = "Plate",
    plate_well      = "Well",
    sample          = "Sample",
    fieldrep_ID     = "Family",
    treatment_group = "Treatment",
    exclude_reason  = "Reason"
  )
  col_names <- col_map[names(col_map) %in% names(excluded_wells)]
  cat(knitr::kable(excluded_wells, col.names = col_names), sep = "\n")
} else {
  cat("No wells are excluded from analysis.\n")
}
```

# 

## Models

Plot raw data.

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = temperature,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ treatment_group) +
  geom_point() +
  scale_colour_manual(values=c("26C"="green4", "11C"="purple2"))+
  geom_line()+
  theme_classic()
```

```{r}
full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             colour = treatment_group,
             group = interaction(plate_ID, plate_well))) +
  facet_wrap(~ temperature) +
  geom_point() +
  geom_line() +
  theme_classic()
```

```{r}
model <- lmer(value.mm ~ timepoint * temperature * treatment_group +
                (1 | plate_ID:plate_well) + (1 | fieldrep_ID),
              data = full_data)

summary(model)
car::Anova(model, type = "II")

qqPlot(residuals(model))
```

Residuals are violating normality. Attempt with cube root transformation.

```{r}
model <- lmer((1 + value.mm)^(1/3) ~ timepoint * temperature * treatment_group +
                (1 | plate_ID:plate_well) + (1 | fieldrep_ID),
              data = full_data)

summary(model)
anova(model)

qqPlot(residuals(model))
```

View posthoc effects.

```{r}
emm <- emmeans(model, ~ treatment_group | temperature | timepoint)
pairs(emm)

emm <- emmeans(model, ~ treatment_group | timepoint)
pairs(emm)

emm <- emmeans(model, ~ timepoint | treatment_group)
pairs(emm)

emm <- emmeans(model, ~ temperature | timepoint)
pairs(emm)
```

Treated and control rates are different at 36C and 40C.

## Plotting

Plot with individual points with geom smooth lines.

```{r}
plot1 <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = treatment_group, fill = treatment_group)) +
  facet_grid(~ fieldrep_ID * plate_rep) +
  geom_point(alpha = 0.5) +
  stat_smooth(aes(group = treatment_group)) +
  theme_classic() +
  xlab("Hour")
plot1

plot1a <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = treatment_group, fill = treatment_group,
             group = treatment_group)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "loess") +
  theme_classic() +
  xlab("Hour")
plot1a

ggsave(plot1a, filename = "output/AP_cockles_SP1_pink/figures/ah/metabolism_treatment.png", width = 7, height = 5)

plot1b <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = fieldrep_ID, fill = fieldrep_ID)) +
  facet_grid(~ temperature) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = fieldrep_ID)) +
  theme_classic() +
  xlab("Hour")
plot1b


plot1c <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = treatment_group, fill = treatment_group)) +
  facet_grid(~ temperature) +
  geom_smooth(aes(group = treatment_group)) +
  theme_classic() +
  theme(axis.text.y  = element_text(size = 17),
        axis.text.x  = element_text(size = 17),
        axis.title.x = element_text(size = 17),
        axis.title.y = element_text(size = 17)) +
  ylab("Metabolic Rate") +
  xlab("Hour")
plot1c

ggsave(plot1c, filename = "output/AP_cockles_SP1_pink/figures/ah/metabolism_treatment_by_temp.png", width = 7, height = 4)

plot1d <- full_data %>%
  ggplot(aes(x = timepoint, y = value.mm,
             color = temperature, fill = temperature, group = temperature)) +
  facet_grid(~ treatment_group) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "loess") +
  theme_classic() +
  xlab("Hour")
plot1d

ggsave(plot1d, filename = "output/AP_cockles_SP1_pink/figures/ah/metabolism_temperature_by_treatment.png", width = 8, height = 5)
```

# Metabolic rates across thermal range

Calculate area under the curve

```{r}
library(pracma)

head(full_data)

auc_data <- full_data %>%
  mutate(time = as.numeric(timepoint)) %>%
  group_by(plate_ID, plate_well, temperature, treatment_group, fieldrep_ID, plate_rep) %>%
  summarise(AUC = trapz(time, value.mm), .groups = "drop") %>%
  mutate(temperature_num = as.numeric(temperature))

head(auc_data)
```

```{r}
hist(auc_data$AUC)
```

Plot total metabolic activity for each treatment over the temperatures.

```{r}
plot5 <- ggplot(auc_data, aes(x = temperature, y = AUC,
                              color = treatment_group, fill = treatment_group,
                              group = treatment_group)) +
  geom_point(alpha = 0.6, position = position_jitter(height = 0.01)) +
  stat_smooth(aes(group = treatment_group)) +
  labs(title = "A", y = "AUC", x = "Temperature") +
  theme_classic(base_size = 16)
plot5


```

Analyze effect of temperature on AUC.

```{r}
tpc_model <- lmer(log(1 + AUC) ~ temperature * treatment_group +
                    (1 | fieldrep_ID) + (1 | plate_ID),
                  data = auc_data)

summary(tpc_model)
anova(tpc_model, type = "II")
qqPlot(residuals(tpc_model))

plot(tpc_model, which = 1)
plot(tpc_model, which = 3)
```

Temperature x treatment is significant

```{r}
emm <- emmeans(tpc_model, ~ treatment_group | temperature)
pairs(emm)
```

Generate a model prediction plot.

```{r}
# Create a grid of temperature and treatment_group values
pred_grid <- expand.grid(
  temperature    = levels(as.factor(auc_data$temperature)),
  treatment_group = levels(as.factor(auc_data$treatment_group)),
  fieldrep_ID    = NA   # ignore random effect
)

pred_grid$treatment_group <- factor(pred_grid$treatment_group,
                                    levels = levels(as.factor(auc_data$treatment_group)))
pred_grid$temperature <- factor(pred_grid$temperature,
                                levels = levels(as.factor(auc_data$temperature)))

# Predicted values (log scale)
pred_grid$logAUC_pred <- predict(tpc_model, newdata = pred_grid, re.form = NA)

# Back-transform to original AUC scale
pred_grid$AUC_pred <- exp(pred_grid$logAUC_pred) - 1

plot5 <- ggplot() +
  geom_point(data = auc_data,
             aes(x = temperature, y = AUC, color = treatment_group),
             alpha = 0.3, position = position_jitter(height = 0.01)) +
  geom_line(data = pred_grid,
            aes(x = temperature, y = AUC_pred,
                color = treatment_group, group = treatment_group),
            linewidth = 1.5) +
  labs(title = "A", y = "AUC", x = "Temperature") +
  theme_classic(base_size = 16)
plot5

ggsave(plot5, filename = "output/AP_cockles_SP1_pink/figures/ah/auc-temperatures.png", width = 7, height = 5)
```

Calculate percent difference.

```{r}
auc_data %>%
  group_by(temperature, treatment_group) %>%
  summarise(mean = mean(AUC, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = treatment_group, values_from = mean)
```
