---
title: "Clam resazurin metabolic rate analysis"
author: 
date: '2026'
params:
  data_dir: "~/Documents/GitHub/rphil_fieldprime/data/WC_manila_res_final"
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes resazurin assays for clam samples from BioTek Synergy HTX exports (24-well plate layout). Blank wells are located at A1, B4, C2, D5; all remaining wells contain clam samples.

Samples span four treatment groups, each run at four exposure temperatures, tracked over multiple timepoints. Files follow the naming scheme `WC-[treatment code][replicate #]-[temp]-[timepoint]`, e.g. `WC-CC1-30-T2` is treatment group CC, replicate 1, exposed to 30°C, timepoint 2. Figures are generated separately for each temperature so that treatment differences within a temperature are easy to compare.

# Set up

Set up workspace, set options, and load required packages.

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries.

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(tools)
```

`knitr::knit()` does not populate `params`; `rmarkdown::render()` does. When knitting outside `render()`, use the same default as `params:` in the YAML header above.

```{r include=FALSE}
if (!exists("params", inherits = FALSE)) {
  params <- list(data_dir = "~/Documents/GitHub/rphil_fieldprime/data/WC_manila_res_final")
}
```

# Define plate layout

Wells containing blanks (all other wells on the 24-well plate are clam samples).

```{r}
blank_wells <- c("A1", "A6", "B1", "B6", "C1","C6", "D1" , "D6")
```

# Load data

## Parse plate reader files

Define a function to parse BioTek Synergy HTX .txt files. Each file may contain one or more plate readings separated by a "Results" header. For files with multiple readings the fluorescence values are averaged across readings. Unlike a design where only columns 1–2 are used, this parses every well on the plate (e.g. columns 1–6 for a 24-well plate) since blank wells are scattered across the layout rather than confined to one column.

```{r}
parse_plate_file <- function(file) {
  lines <- readLines(file, warn = FALSE)
  
  # Find all lines that contain only "Results"
  results_idx <- which(trimws(lines) == "Results")
  
  if (length(results_idx) == 0) {
    stop(paste("No 'Results' section found in file:", file))
  }
  
  readings_list <- list()
  
  for (idx in results_idx) {
    # Line idx   = "Results"
    # Line idx+1 = column header (tab-separated well columns; 6 cols for 24-well, 8 for 48-well)
    # Data rows: A–D (24-well) or A–F (48-well), then blank line or end of section
    j <- idx + 2
    data_lines <- character()
    while (j <= length(lines)) {
      line <- lines[j]
      if (nchar(trimws(line)) == 0) break
      if (!grepl("^[A-Fa-f]\\t", line)) break
      data_lines <- c(data_lines, line)
      j <- j + 1
    }

    plate_data <- lapply(data_lines, function(row) {
      parts <- unlist(strsplit(row, "\t"))
      row_letter <- trimws(parts[1])
      # Last field is filter label (e.g. 528/20,590/20); preceding fields are well values
      n <- length(parts)
      value_parts <- if (n >= 3) parts[seq.int(2, n - 1)] else parts[-1]
      values <- suppressWarnings(as.numeric(value_parts))
      data.frame(
        well  = paste0(row_letter, seq_along(values)),
        fluor = values,
        stringsAsFactors = FALSE
      )
    })
    
    readings_list[[length(readings_list) + 1]] <- bind_rows(plate_data)
  }
  
  # If multiple readings exist, average per well
  if (length(readings_list) > 1) {
    bind_rows(readings_list, .id = "reading") %>%
      group_by(well) %>%
      summarise(fluor = mean(fluor, na.rm = TRUE), .groups = "drop")
  } else {
    readings_list[[1]]
  }
}
```

Read and combine all plate files from `params$data_dir`. Filenames follow the pattern `WC-[treatment code][replicate #]-[temp]-[timepoint].txt`, e.g. `WC-CC1-30-T2.txt` is treatment group CC, replicate 1, exposed to 30°C, timepoint 2. A given treatment/replicate/temperature combination represents one physical plate that is read repeatedly over time, so those three fields together (`plate_id`) identify the plate, while `timepoint` varies across files for that plate. To reuse this script for a new run, just change `data_dir` in the YAML header (or pass it via `rmarkdown::render(..., params = list(data_dir = "..."))`) — everything else, including where figures are saved, follows automatically.

```{r}
folder_path <- path.expand(params$data_dir)

file_list <- list.files(path = folder_path, pattern = "\\.txt$",
                        full.names = TRUE, ignore.case = TRUE)

if (length(file_list) == 0) {
  stop("No .txt files found in ", folder_path, ". Check params$data_dir.")
}

# Mirror the data folder's path under a sibling "figures" directory,
# e.g. .../data/test1/ctl -> .../figures/test1/ctl
fig_dir <- sub("/data/", "/figures/", folder_path, fixed = TRUE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

data_list <- list()

# WC-[treatment code][replicate #]-[temp]-[timepoint], e.g. WC-CC1-30-T2
filename_pattern <- "^wc-([a-z]+)([0-9]+)-([0-9.]+)-t([0-9.]+)$"

for (file in file_list) {
  file_name <- file_path_sans_ext(basename(file))
  fn_lower  <- tolower(file_name)

  m <- str_match(fn_lower, filename_pattern)

  if (is.na(m[, 1])) {
    stop("Filename '", file_name, "' does not match the expected pattern ",
         "WC-[treatment][replicate]-[temp]-[timepoint], e.g. WC-CC1-30-T2.")
  }

  treatment <- toupper(m[, 2])
  replicate <- as.integer(m[, 3])
  temp      <- as.numeric(m[, 4])
  timepoint <- as.numeric(m[, 5])
  plate_id  <- paste(treatment, replicate, temp, sep = "_")

  parsed <- parse_plate_file(file) %>%
    mutate(
      treatment = treatment,
      replicate = replicate,
      temp      = temp,
      timepoint = timepoint,
      plate_id  = plate_id,
      well_type = if_else(well %in% blank_wells, "blank", "sample")
    )
  
  data_list[[file_name]] <- parsed
}

combined_data <- bind_rows(data_list)

head(combined_data)
```

## Normalize to T0

Divide each well's fluorescence by the T0 (baseline) value to express fluorescence as fold-change relative to the start of the assay. Normalization is done per physical plate (`plate_id` = treatment + replicate + temp), since the same plate is read repeatedly across timepoints.

```{r}
combined_data <- combined_data %>%
  group_by(plate_id, well) %>%
  arrange(timepoint, .by_group = TRUE) %>%
  mutate(fluor_norm = fluor / first(fluor)) %>%
  ungroup()

head(combined_data)
```

## Blank correction

Calculate the mean normalized blank fluorescence per plate and timepoint (using only wells A1, B4, C2, D5), then subtract it from each normalized sample well's fluorescence (adding 1 to keep values centred at 1 at T0).

```{r}
blanks_summary <- combined_data %>%
  filter(well_type == "blank") %>%
  group_by(plate_id, timepoint) %>%
  summarise(mean_blank_norm = mean(fluor_norm, na.rm = TRUE), .groups = "drop")

combined_data <- left_join(combined_data, blanks_summary, by = c("plate_id", "timepoint")) %>%
  mutate(fluorescence_corr = fluor_norm - mean_blank_norm + 1)

head(combined_data)
```

# Plotting

Blank-corrected plots below use sample wells only, since blank wells are the correction reference rather than data of interest. Figures are generated separately for **each exposure temperature**, so that differences between the four treatment groups within a temperature are easy to see. Each temperature section below produces the same five plot types: raw fluorescence, normalized fluorescence, blank-corrected fluorescence, mean ± SE by treatment, and individual well trajectories faceted by treatment.

```{r}
sample_data <- combined_data %>% filter(well_type == "sample")

temps <- sort(unique(combined_data$temp))
temps
```

## Figures by temperature

```{r, results = "asis", fig.width = 7, fig.height = 4}
for (temp_val in temps) {

  temp_label <- paste0(temp_val, "\u00B0C")
  cat(sprintf("\n\n### %s\n\n", temp_label))

  temp_all    <- combined_data %>% filter(temp == temp_val)
  temp_sample <- sample_data   %>% filter(temp == temp_val)

  # --- Raw fluorescence over time -----------------------------------------
  plot_raw <- temp_all %>%
    ggplot(aes(x = timepoint, y = fluor, colour = treatment,
               group = interaction(treatment, replicate, well))) +
    geom_point() +
    geom_line() +
    labs(
      x      = "Timepoint",
      y      = "Raw Fluorescence",
      colour = "Treatment",
      title  = paste("Raw fluorescence -", temp_label)
    ) +
    theme_classic()
  print(plot_raw)
  ggsave(plot_raw,
         filename = file.path(fig_dir, sprintf("raw_fluorescence_%sC.png", temp_val)),
         width = 6, height = 4)

  # --- Normalized fluorescence over time ----------------------------------
  plot_norm <- temp_all %>%
    ggplot(aes(x = timepoint, y = fluor_norm, colour = treatment,
               group = interaction(treatment, replicate, well))) +
    geom_point() +
    geom_line() +
    labs(
      x      = "Timepoint",
      y      = "Fluorescence (fold change from T0)",
      colour = "Treatment",
      title  = paste("Normalized fluorescence -", temp_label)
    ) +
    theme_classic()
  print(plot_norm)
  ggsave(plot_norm,
         filename = file.path(fig_dir, sprintf("normalized_fluorescence_%sC.png", temp_val)),
         width = 6, height = 4)

  # --- Blank-corrected fluorescence over time -----------------------------
  plot_corr <- temp_sample %>%
    ggplot(aes(x = timepoint, y = fluorescence_corr, colour = treatment,
               group = interaction(treatment, replicate, well))) +
    geom_point() +
    geom_line() +
    labs(
      x      = "Timepoint",
      y      = "Blank-corrected Fluorescence",
      colour = "Treatment",
      title  = paste("Blank-corrected fluorescence -", temp_label)
    ) +
    theme_classic()
  print(plot_corr)
  ggsave(plot_corr,
         filename = file.path(fig_dir, sprintf("corrected_fluorescence_%sC.png", temp_val)),
         width = 6, height = 4)

  # --- Mean metabolic activity over time, by treatment --------------------
  summary_temp <- temp_sample %>%
    group_by(treatment, timepoint) %>%
    summarise(
      mean_fluor = mean(fluorescence_corr, na.rm = TRUE),
      se_fluor   = sd(fluorescence_corr,   na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    )

  plot_mean <- summary_temp %>%
    ggplot(aes(x = timepoint, y = mean_fluor,
               colour = treatment, fill = treatment)) +
    geom_ribbon(aes(ymin = mean_fluor - se_fluor, ymax = mean_fluor + se_fluor),
                alpha = 0.2, colour = NA) +
    geom_point() +
    geom_line() +
    labs(
      x      = "Timepoint",
      y      = "Blank-corrected Fluorescence (mean \u00B1 SE)",
      colour = "Treatment",
      fill   = "Treatment",
      title  = paste("Mean metabolic activity -", temp_label)
    ) +
    theme_classic()
  print(plot_mean)
  ggsave(plot_mean,
         filename = file.path(fig_dir, sprintf("mean_fluorescence_%sC.png", temp_val)),
         width = 6, height = 4)

  # --- Individual well trajectories faceted by treatment ------------------
  plot_facet <- temp_sample %>%
    ggplot(aes(x = timepoint, y = fluorescence_corr,
               colour = factor(replicate), group = interaction(replicate, well))) +
    facet_wrap(~treatment) +
    geom_point() +
    geom_line() +
    labs(
      x      = "Timepoint",
      y      = "Blank-corrected Fluorescence",
      colour = "Replicate",
      title  = paste("Individual wells by treatment -", temp_label)
    ) +
    theme_classic()
  print(plot_facet)
  ggsave(plot_facet,
         filename = file.path(fig_dir, sprintf("individual_wells_%sC.png", temp_val)),
         width = 8, height = 4)
}
```
