---
title: "Retrieval Rate Analysis: 2x2 Factorial Priming Experiment"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This script analyses retrieval rates from a **2×2 factorial priming experiment** with two binary factors:

- **Heat priming** (No / Yes)
- **Immune priming** (No / Yes)

The response variable is the number of individuals retrieved out of those deployed (binomial outcome), modelled using a **binomial GLM** (logit link).

---

# 1. Load Packages

```{r packages}
library(tidyverse)
library(emmeans)
library(car)        # Anova() with Type III SS
library(ggplot2)
```

---

# 2. Enter Data

```{r data}
data <- data.frame(
  priming_treatment = c("Control", "Heat", "Immune", "Combination"),
  heat              = c(0, 1, 0, 1),   # 0 = No heat priming, 1 = Heat priming
  immune            = c(0, 0, 1, 1),   # 0 = No immune priming, 1 = Immune priming
  retrieved         = c(45, 42, 32, 39),
  deployed          = c(120, 120, 120, 120)
) %>%
  mutate(
    rate             = retrieved / deployed,
    heat_label       = factor(ifelse(heat   == 1, "Heat",   "No Heat"),
                              levels = c("No Heat", "Heat")),
    immune_label     = factor(ifelse(immune == 1, "Immune", "No Immune"),
                              levels = c("No Immune", "Immune")),
    priming_treatment = factor(priming_treatment,
                               levels = c("Control", "Heat", "Immune", "Combination"))
  )

# Preview
knitr::kable(
  data %>% select(priming_treatment, heat_label, immune_label, retrieved, deployed, rate),
  digits  = 3,
  caption = "Retrieval rates by priming treatment"
)
```

---

# 3. Binomial GLM

## 3.1 Fit Full Factorial Model

```{r model-full}
model_full <- glm(
  cbind(retrieved, deployed - retrieved) ~ heat * immune,
  data   = data,
  family = binomial(link = "logit")
)

summary(model_full)
```

> **Note:** With only one observation per cell (no within-cell replication), the saturated model has **zero residual degrees of freedom**. The interaction term is confounded with residual error — interpretation below accounts for this.

## 3.2 Likelihood Ratio Tests (Type II)

```{r anova-lrt}
anova(model_full, test = "Chisq")
```

## 3.3 Main-Effects-Only Model (assumes no interaction)

If the interaction is not meaningful or cannot be estimated, fit the additive model:

```{r model-additive}
model_add <- glm(
  cbind(retrieved, deployed - retrieved) ~ heat + immune,
  data   = data,
  family = binomial(link = "logit")
)

summary(model_add)
anova(model_add, test = "Chisq")
```

## 3.4 Model Comparison

```{r model-comparison}
anova(model_add, model_full, test = "Chisq")
```

---

# 4. Estimated Marginal Means & Pairwise Contrasts

```{r emmeans}
# Estimated marginal means on the response (probability) scale
emm <- emmeans(model_add, ~ heat * immune, type = "response")
emm

# All pairwise comparisons (Tukey-adjusted)
pairs(emm, adjust = "tukey")
```

```{r contrasts-table}
contrast_df <- as.data.frame(pairs(emm, adjust = "tukey"))
knitr::kable(contrast_df, digits = 4, caption = "Pairwise contrasts (Tukey-adjusted)")
```

---

# 5. Model Diagnostics

```{r diagnostics, fig.width=8, fig.height=6}
par(mfrow = c(2, 2))
plot(model_add, which = 1:4)
par(mfrow = c(1, 1))
```

---

# 6. Visualisation

## 6.1 Interaction Plot

```{r interaction-plot, fig.width=6, fig.height=4}
# Get predicted probabilities from the additive model
pred_data <- data %>%
  mutate(
    predicted = predict(model_add, type = "response")
  )

ggplot(pred_data, aes(x = heat_label, y = rate,
                      colour = immune_label, group = immune_label)) +
  geom_line(aes(y = predicted), linetype = "dashed", linewidth = 0.8) +
  geom_point(size = 4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.5)) +
  scale_colour_manual(values = c("No Immune" = "#2196F3", "Immune" = "#F44336")) +
  labs(
    title    = "Retrieval Rate by Priming Treatment",
    subtitle = "Points = observed; dashed lines = additive model predictions",
    x        = "Heat Priming",
    y        = "Retrieval Rate",
    colour   = "Immune Priming"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## 6.2 Bar Chart with Observed Rates

```{r bar-chart, fig.width=6, fig.height=4}
ggplot(data, aes(x = priming_treatment, y = rate, fill = priming_treatment)) +
  geom_col(width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(retrieved, "/", deployed)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.5)) +
  scale_fill_manual(values = c(
    "Control"     = "#90CAF9",
    "Heat"        = "#EF9A9A",
    "Immune"      = "#A5D6A7",
    "Combination" = "#CE93D8"
  )) +
  labs(
    title = "Observed Retrieval Rates by Treatment",
    x     = "Priming Treatment",
    y     = "Retrieval Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# 7. Results Summary

```{r results-table}
results <- data %>%
  mutate(
    predicted_rate = predict(model_add, type = "response"),
    pct_retrieved  = paste0(round(rate * 100, 1), "%")
  ) %>%
  select(priming_treatment, heat_label, immune_label,
         retrieved, deployed, pct_retrieved, predicted_rate)

knitr::kable(results, digits = 3,
             caption = "Observed and model-predicted retrieval rates")
```

---

# 8. Notes on Replication

> **Important:** This analysis has **one observation per treatment cell**. Without within-cell replication (e.g. multiple tanks or batches per treatment), it is not possible to:
>
> - Independently estimate the interaction term and residual error
> - Reliably assess overdispersion
> - Compute within-group variance
>
> If multiple tanks/replicates exist per treatment, restructure the data to include a replicate ID and consider a **mixed-effects binomial model** (`lme4::glmer`) with tank as a random effect:
>
> ```r
> library(lme4)
> model_mixed <- glmer(
>   cbind(retrieved, deployed - retrieved) ~ heat * immune + (1 | tank),
>   data   = data_rep,
>   family = binomial
> )
> ```

---

# Session Info

```{r session-info}
sessionInfo()
```
