---
title: "Manila Clam Survival Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options:
  chunk_output_type: console
---

This script analyzes survival assay data for lab-reared Manila clams in a factorial immune priming × heat priming experiment. Mortality (0 = alive, 1 = dead) is modeled over time using mixed-effects binomial logistic regression, with silo, field replicate, and assay plate included as random effects.

# Setup

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(lme4)
library(car)
library(emmeans)
library(DHARMa)
library(ggplot2)
library(cowplot)
library(lmerTest)
```

# Load data

```{r load}
# ── CONFIG ────────────────────────────────────────────────────────────────────
# Point this at your transformed mortality CSV (output of transform_mortality.R).
# If running multiple assay temperatures, either combine files before loading
# or re-run this script once per file.
DATA_FILE <- "data/LR_manila-40C_mortality.csv"
# ─────────────────────────────────────────────────────────────────────────────

surv <- read_csv(DATA_FILE)

# Confirm structure
glimpse(surv)
```

# Wrangle

Pivot timepoint columns to long format so each row is one clam × one timepoint.

```{r wrangle}
# Identify timepoint columns (anything ending in "HR")
tp_cols <- names(surv)[grepl("HR$", names(surv), ignore.case = TRUE)]

surv_long <- surv %>%
  pivot_longer(
    cols      = all_of(tp_cols),
    names_to  = "timepoint",
    values_to = "mortality"
  ) %>%
  mutate(
    # Parse numeric hours from labels like "4HR", "29.5HR"
    hours = as.numeric(sub("HR$", "", timepoint, ignore.case = TRUE)),

    # Ensure factor variables are correctly typed
    Immune_Priming    = factor(Immune_Priming,    levels = c("No", "Yes")),
    Heat_Priming      = factor(Heat_Priming,      levels = c("No", "Yes")),
    Silo_Replicate    = factor(Silo_Replicate),
    Field_Replicate   = factor(Field_Replicate),
    Assay_Replicate   = factor(Assay_Replicate),
    timepoint         = factor(timepoint, levels = tp_cols),  # preserve file order

    # Mortality must be integer 0/1 for binomial family
    mortality = as.integer(mortality)
  )

# Quick sanity check
surv_long %>%
  count(Immune_Priming, Heat_Priming, Silo_Replicate, Field_Replicate, Assay_Replicate) %>%
  print(n = Inf)
```

# Analysis: full time series

Mixed-effects binomial logistic regression across all timepoints.

Fixed effects: `Immune_Priming * Heat_Priming * hours`\
Random effects: `Silo_Replicate`, `Field_Replicate` (nested in silo), `Assay_Replicate` (plate)

```{r model-timeseries}
model_ts <- glmer(
  mortality ~ Immune_Priming * Heat_Priming * hours +
    (1 | Silo_Replicate) +
    (1 | Silo_Replicate:Field_Replicate) +
    (1 | Assay_Replicate),
  family = binomial(link = "logit"),
  data   = surv_long,
  control = glmerControl(optimizer = "bobyqa",
                         optCtrl   = list(maxfun = 2e5))
)

summary(model_ts)
Anova(model_ts, type = "II")
```

Check significance of random effects.

```{r rand-effects-ts}
model_no_silo  <- update(model_ts, . ~ . - (1 | Silo_Replicate))
model_no_field <- update(model_ts, . ~ . - (1 | Silo_Replicate:Field_Replicate))
model_no_assay <- update(model_ts, . ~ . - (1 | Assay_Replicate))

anova(model_ts, model_no_silo)
anova(model_ts, model_no_field)
anova(model_ts, model_no_assay)

VarCorr(model_ts)
```

Post-hoc: pairwise comparisons of treatment combinations at each timepoint.

```{r posthoc-ts}
emm_ts <- emmeans(model_ts, ~ Immune_Priming * Heat_Priming | hours, type = "response")
pairs(emm_ts)
```

## Plot: predicted mortality over time

```{r plot-timeseries}
surv_long$predicted_mortality <- predict(model_ts, type = "response")

plot_ts <- ggplot(surv_long,
                  aes(x     = hours,
                      y     = predicted_mortality,
                      color = interaction(Immune_Priming, Heat_Priming),
                      fill  = interaction(Immune_Priming, Heat_Priming),
                      group = interaction(Immune_Priming, Heat_Priming))) +
  geom_point(aes(y = mortality), alpha = 0.3,
             position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = TRUE, alpha = 0.15) +
  scale_colour_manual(
    values = c("No.No"   = "grey40",
               "Yes.No"  = "steelblue",
               "No.Yes"  = "coral2",
               "Yes.Yes" = "mediumpurple3"),
    labels = c("No.No"   = "Control (CC)",
               "Yes.No"  = "Immune only (CP)",
               "No.Yes"  = "Heat only (EE)",
               "Yes.Yes" = "Immune + Heat (EP)"),
    name = "Treatment"
  ) +
  scale_fill_manual(
    values = c("No.No"   = "grey40",
               "Yes.No"  = "steelblue",
               "No.Yes"  = "coral2",
               "Yes.Yes" = "mediumpurple3"),
    labels = c("No.No"   = "Control (CC)",
               "Yes.No"  = "Immune only (CP)",
               "No.Yes"  = "Heat only (EE)",
               "Yes.Yes" = "Immune + Heat (EP)"),
    name = "Treatment"
  ) +
  labs(
    title = paste("Predicted Mortality over Time —", unique(surv_long$Assay_Temp)),
    x     = "Hours",
    y     = "Probability of Mortality"
  ) +
  theme_classic(base_size = 13); plot_ts

ggsave(plot_ts,
       filename = "figures/survival_timeseries.png",
       width = 8, height = 5)
```

# Analysis: final timepoint only

Isolate the last recorded timepoint and refit a simpler model for a clean endpoint comparison.

```{r model-final}
final_tp <- max(surv_long$hours)

surv_final <- surv_long %>% filter(hours == final_tp)

model_final <- glmer(
  mortality ~ Immune_Priming * Heat_Priming +
    (1 | Silo_Replicate) +
    (1 | Silo_Replicate:Field_Replicate) +
    (1 | Assay_Replicate),
  family = binomial(link = "logit"),
  data   = surv_final
)

summary(model_final)
Anova(model_final, type = "II")
```

## Overdispersion & zero-inflation checks

```{r diagnostics}
sim <- simulateResiduals(model_final)
testDispersion(sim)
testZeroInflation(sim)
```

## Random effects significance

```{r rand-effects-final}
rand(model_final)
```

## Post-hoc comparisons

```{r posthoc-final}
# Treatment combinations within each priming factor
emm_immune <- emmeans(model_final, ~ Immune_Priming | Heat_Priming, type = "response")
pairs(emm_immune)

emm_heat <- emmeans(model_final, ~ Heat_Priming | Immune_Priming, type = "response")
pairs(emm_heat)

# All four treatment combinations
emm_all <- emmeans(model_final, ~ Immune_Priming * Heat_Priming, type = "response")
pairs(emm_all)
emm_df <- as.data.frame(emm_all)
emm_df
```

## Plot: predicted mortality at final timepoint

```{r plot-final}
surv_final$predicted_mortality <- predict(model_final, type = "response")

plot_final <- ggplot(surv_final,
                     aes(x     = interaction(Immune_Priming, Heat_Priming),
                         y     = predicted_mortality,
                         color = interaction(Immune_Priming, Heat_Priming),
                         fill  = interaction(Immune_Priming, Heat_Priming))) +
  geom_jitter(aes(y = mortality), width = 0.2, height = 0.03, alpha = 0.4) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  scale_colour_manual(
    values = c("No.No"   = "grey40",
               "Yes.No"  = "steelblue",
               "No.Yes"  = "coral2",
               "Yes.Yes" = "mediumpurple3"),
    labels = c("No.No"   = "Control (CC)",
               "Yes.No"  = "Immune only (CP)",
               "No.Yes"  = "Heat only (EE)",
               "Yes.Yes" = "Immune + Heat (EP)"),
    name = "Treatment"
  ) +
  scale_fill_manual(
    values = c("No.No"   = "grey40",
               "Yes.No"  = "steelblue",
               "No.Yes"  = "coral2",
               "Yes.Yes" = "mediumpurple3"),
    labels = c("No.No"   = "Control (CC)",
               "Yes.No"  = "Immune only (CP)",
               "No.Yes"  = "Heat only (EE)",
               "Yes.Yes" = "Immune + Heat (EP)"),
    name = "Treatment"
  ) +
  scale_x_discrete(
    labels = c("No.No"   = "Control\n(CC)",
               "Yes.No"  = "Immune only\n(CP)",
               "No.Yes"  = "Heat only\n(EE)",
               "Yes.Yes" = "Immune + Heat\n(EP)")
  ) +
  labs(
    title = paste0("Predicted Mortality at ", final_tp, " h — ", unique(surv_long$Assay_Temp)),
    x     = "Treatment",
    y     = "Probability of Mortality"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none"); plot_final

ggsave(plot_final,
       filename = "figures/survival_final_timepoint.png",
       width = 7, height = 5)
```
